---
title: "Mapping VPL Data"
author: "Alexander Ondrus"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals

Using pre-processed data containing time stamped _Vehicle Position Logs_ (VPLs), I want to create animated maps that show the following:

1. The position of each bus in the city on 5 minute intervals
2. The average prediction for derivation from the schedule (early or late) for all busses in each community, by hour
3. The average speed for all busses passing through each neighbourhood, by hour

## Loading Libraries and Data

First, I load in the vehicle position logs along with libraries for aggregating and manipulating data, working with dates, and working with geospatial information. I also load in a shapefile of Calgary community boundaries from the City of Calgary Open Data Catalogue [here](https://data.calgary.ca/Base-Maps/Community-Boundaries/ab7m-fwn6).

```{r loading libraries and data}
library(tidyverse)
library(lubridate)
library(rgdal)
library(rgeos)
library(maptools)
library(gifski)
library(gganimate)
library(transformr)

vpl_15052019 <- readRDS("vpl_15052019.rds")
community_boundaries <- readOGR("geo_export_e135fc9f-0eec-429a-b94b-14836af732d4.shp")
community_bounds_df <- fortify(community_boundaries,
                               region = "name")
```

## First Map: Vehicle Locations

For this map, I need to have a data frame with the time stamps synchronized so that I can plot them on the same frame. I do this by rounding each time stamp to the nearest 5 minutes, and then aggregating the vehicle positions by taking the mean of the vehicle positions over that 5 minute interval.

```{r synchronizing time stamps}
vpl_5_min_intervals <- vpl_15052019
vpl_5_min_intervals$vehicle_position_date_time <-
  round_date(vpl_5_min_intervals$vehicle_position_date_time,
             unit = "5 minutes")
vpl_5_min_intervals <- group_by(vpl_5_min_intervals,
                                vehicle_position_date_time,
                                vehicle_id) %>% 
  summarise(longitude = mean(longitude),
            latitude = mean(latitude)) %>% 
  arrange(vehicle_position_date_time, vehicle_id)
```

Now I generate the animated map. Note that the `transition_states()` argument only applies to the layers using the data passed in to `ggplot()`. Also note that  `coord_map()` uses the mercator projection by default. 

```{r generate animated vehicle location map}
p <- ggplot(vpl_5_min_intervals,
            aes(x = longitude,
                y = latitude)) +
  geom_point(aes(group = vehicle_id),
             colour = "red") +
  labs(title = "Position of YYC Transit Vehicles",
       subtitle = "{closest_state}",
       caption = "Data: City of Calgary Open Data\nCreated by Alexander Ondrus") +
  geom_polygon(data = community_bounds_df,
               aes(x = long,
                   y = lat,
                   group = group),
               fill = NA,
               colour = "grey") +
  theme_void() +
  theme(plot.margin = margin(5,5,5,5)) +
  coord_map() +
  transition_states(vehicle_position_date_time,
                    transition_length = 4,
                    state_length = 1)
  
#animate(p, nframes = 1440, duration = 60)
```

## Second Map: Early or Late?

For the next map, I want to colour each community by the mean predicted deviation for all of the transit vehicle that cross through that community in a given 15-minute span.

```{r calculate mean deviation for communities}
vpl_deviation <- vpl_15052019 %>% 
  mutate(time_rounded_1h = round_date(vehicle_position_date_time,
                                        unit = "1 hour")) %>% 
  group_by(time_rounded_1h,
           community) %>% 
  summarise(med_deviation = median(predicted_deviation))

vpl_deviation$deviation <- cut_interval(vpl_deviation$med_deviation, 
                                        n = 7,
                                        dig.lab = 2,
                                        ordered_result = TRUE)
```

Note that `vpl_deviation` does not contain `NA` values for time-community combinations that are not present, they are simply not listed. I need the NA values to be present in order for them to be represented on the choropleth.

To make this possible, I will take the Cartesian product of the `community_bounds_df` with the unique time values before attaching the mean deviations with a `left_join()`.

```{r create df with boundaries and choropleth}
times <- vpl_deviation %>% 
  arrange(time_rounded_1h) %>% 
  select(time_rounded_1h) %>% 
  distinct()

community_bounds_w_times <- crossing(community_bounds_df,
                                     times) %>% 
  arrange(order, time_rounded_1h)

community_bounds_w_deviation <- left_join(community_bounds_w_times,
                                          vpl_deviation,
                                          by = c("id" = "community",
                                                 "time_rounded_1h" = "time_rounded_1h"))
```

I am now in a position to map the data using `geom_polygon` with the fill mapped to the mean deviation.

```{r construct map 2}
map_2 <- ggplot(community_bounds_w_deviation,
                aes(x = long,
                    y = lat,
                    group = group)) +
  geom_polygon(aes(fill = deviation),
               colour = "white") +
  coord_map() +
  theme_void() +
  theme(plot.margin = margin(5,5,5,5)) +
  transition_time(time_rounded_1h) +
  scale_fill_brewer(type = "div",
                    palette = "Spectral") + 
  labs(title = "Will YYC Transit Vehicles be Early or Late?",
       subtitle = "Median predicted deviation from schedule for all vehicles\nin each neighbourhood at times closest to {frame_time}",
       caption = "Data: City of Calgary Open Data Catalogue\nCreated by: Alexander Ondrus",
       fill = "Predicted Deviation")

#animate(map_2, nframes = 25, duration = 30)
```

